--- 
title_meta  : Chapter 2
title       : Data the Tidy Way
description : "In this chapter, you will learn some of the fundamentals of taking raw data and cleaning it to form tidy data, the fundamental currency of data science and visualization tools such as ggplot2."

--- type:NormalExercise xp:0 skills:1 key:3db54e3447
## Intro to Data Science

Now that you've been introduced to the fundamentals of programming in R, the next step is to learning how to data science in R. Here, we'll present some of the concepts we wish we would have learned when we first started, and go through some example data on the way. (see hint)

By the end of this chapter, you should be familiar with:

- understanding how and why R is useful for data science
- explain the advantages and concepts of tidy data
- familiarity with modern R conventions that make it a powerful tool for data science

*** =instructions
Simply click submit to continue!
*** =hint
Did you know: the R language is a dialect of S which was designed in the 1980s and has been in widespread use in the statistical community since. 

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
## Hiding within those mounds of data is knowledge
## that could change the life of a patient,
## or change the world." - Atul Butte

```

*** =Solution
```{r}
## No code on this one, just hit Submit Answer to continue
```

*** =sct
```{r}
success_msg("Let's get started!")
```

--- type:NormalExercise lang:r xp:0 skills:1,4 key:c24bed01fc
## Some (Un)Tidy Data

On to our first concept: **tidy data** (see hint)

Many times when we run into data from the wild, it is often a jumble of messy character strings, confounding variables, missing data, and worse. The process of converting a *raw data* to *tidy data* goes by many names - here we will refer to it as *data munging* or *cleaning*. The goal of this process is singular:

> **"A dataset is a collection of *values*, organized in two ways: every value belongs to a *variable* and *observation*.**
> **A *variable* contains all *values* that measure the same underlying attribute across units.**
> **An *observation* contains all *values* measured on the same unit across attributes.**

To emphasize this concept, take a look at the sample code on the right showing untidy data (presented in two different ways), and print the `untidy_df` and `untidy_df_2` object out (hint: Ctrl/Cmd+Enter to run a line!). How does this dataset run counter to the principle stated above? 


*** =instructions
- Take a moment to ponder how you might reorganize this data, then simply click submit to move forward!


*** =hint
These concepts introduced here lean heavily on the formal treatise by Hadley Wickham, the lead developer of many of the packages we will be using throughout this course. See the [tidy data paper](http://vita.had.co.nz/papers/tidy-data.html) and the accompanying [R code heavy version](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html). Definitely take the time to checkout these resources!


*** =pre_exercise_code
```{r}
```

*** =sample_code
```{r}
## Untidy data example
untidy_df <- data.frame(name = c("John Smith", "Jane Doe", "Mary Johnson"),
                        treatmenta = c(NA, 4, 6), treatmentb = c(18, 1, 7))
untidy_df

# Untidy data example 2
untidy_df_2 <- data.frame(treatment = c("a", "b"), John.Smith = c(NA, 18),
                          Jane.Doe = c(4, 1), Mary.Johnson = c(6, 7))
untidy_df_2
```

*** =solution
```{r}
## Untidy data example
untidy_df <- data.frame(name = c("John Smith", "Jane Doe", "Mary Johnson"),
                        treatmenta = c(NA, 4, 6), treatmentb = c(18, 1, 7))
untidy_df

# Untidy data example 2
untidy_df_2 <- data.frame(treatment = c("a", "b"), John.Smith = c(NA, 18),
                          Jane.Doe = c(4, 1), Mary.Johnson = c(6, 7))
untidy_df_2
```

*** =sct
```{r}
success_msg("Head over to the next exercise!")
```

--- type:NormalExercise lang:r xp:0 skills:1,4 key:489128dafe
## A Tidy Sample

So we just saw what some untidy looks like, with all of its variable and values mixed up. If you took a moment to ponder the format of the previous data (to remind yourself, run the line `untidy_df` and see the output in the console), you took stock of the following.

- There exist three different people with different *names*, who underwent different *treatments*, `treatmenta` and/or `treatmentb`, which resulted in different *values* in each case (with no treatment application (e.g. missing data) represented as `NA`)

Given this, we can come up with three *variables* describing each unique *observation* (application of treatment X to person Y with readout/value Z):

- name
- treatment
- value

Hence, we can rewrite our untidy data into a tidy format! See the sample code for the solution. 

With this in mind, let's move on to a real world dataset 'from the wild'!

*** =instructions
Check out the tidy transformation of our untidy data, and get ready to start coding!

*** =hint
"It is often said that 80% of data analysis is spent on the process of cleaning and preparing
the data (Dasu and Johnson 2003)" - if you haven't already, check out [this paper on tidy data](http://vita.had.co.nz/papers/tidy-data.pdf) for extra credit!

*** =pre_exercise_code
```{r}
untidy_df <- data.frame(name = c("John Smith", "Jane Doe", "Mary Johnson"),
                        treatmenta = c(NA, 4, 6), treatmentb = c(18, 1, 7))
```

*** =sample_code
```{r}
untidy_df # run me to see the untidy version

name <- rep(c("Jane Doe", "John Smith", "Mary Johnson"), each = 2)
treatment <- rep(c("a", "b"), 3)
value <- c(4, 1, NA, 18, 6, 7)

tidy_df <- data.frame(name, treatment, value)
                      
tidy_df
```

*** =solution
```{r}
untidy_df # run me to see the untidy version

name <- rep(c("Jane Doe", "John Smith", "Mary Johnson"), each = 2)
treatment <- rep(c("a", "b"), 3)
value <- c(4, 1, NA, 18, 6, 7)

tidy_df <- data.frame(name, treatment, value)
                      
tidy_df
```

*** =sct
```{r}
success_msg("Head over to the next exercise!")
```

--- type:NormalExercise lang:r xp:100 skills:1,4 key:75bda68f14
## Real (Messy) Data: Income Distribution by Religion

Time for you to take a look at some actual, raw data. Here we have a survey done by Pew Research that examines the relationship between income and religious affiliation. We've already loaded the dataset up for you in the background under `pew`, but if you want to explore it on your own computer, [here is the data](http://stat405.had.co.nz/data/pew.txt).

Recalling chapter 1, check out the dataset on your own, and let's apply our tidy thinking to figure out what we need to change.

*** =instructions
- What do the first six and last six rows of the data look like?
- What are the dimensions? 
- What does a simple (naive) summary of the data give us? 
- What is the structure of the data?

*** =hint
Use the functions you learned from chapter 1 - namely, [`head()`](http://www.rdocumentation.org/packages/utils/functions/head), [`tail()`](http://www.rdocumentation.org/packages/utils/functions/tail), [`dim()`](http://www.rdocumentation.org/packages/base/functions/dim), [`summary()`](http://www.rdocumentation.org/packages/base/functions/summary), and [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) to preview the dataset on hand.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/pew.RData"))
```

*** =sample_code
```{r}
## What do the first six and last six rows of the data look like?


## What are the dimensions?


## What does a simple (naive) summary of the data give us?


## What is the structure of the data?
```

*** =solution
```{r}
## What do the first six and last six rows of the data look like?
head(pew)
tail(pew)

## What are the dimensions?
dim(pew)

## What does a simple (naive) summary of the data give us?
summary(pew)

## What is the structure of the data?
str(pew)
```

*** =sct
```{r}
test_output_contains("head(pew)",
                     incorrect_msg = "Did you look at the first six lines of `pew`")
test_output_contains("tail(pew)",
                     incorrect_msg = "Did you look at the last six lines of `pew`")
test_or(
    test_output_contains("nrow(pew)"),
    test_output_contains("dim(pew)"),
    incorrect_msg = "Did you determine the number of rows in `pew`?")
test_or(
    test_output_contains("ncol(pew)"),
    test_output_contains("dim(pew)"),
    incorrect_msg = "Did you determine the number of columns in `pew`?")
test_output_contains("summary(pew)",
                     incorrect_msg = "Did you use `summary` to summarize the data in `pew`?")
test_output_contains("str(pew)",
                     incorrect_msg = "Did you use `str` to get the structure of the data in `pew`?")
success_msg("If you could answer all those questions, great job. If not, go back and look at the hint to see how you can answer them, then head over to the next exercise!")
```

--- type:NormalExercise lang:r xp:100 skills:1,4 key:426a68171c
## Refactoring Columns Into a Variable

Before we go on, first, I highly recommend checking out the [Data Wrangling with dplyr and tidyr Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) by RStudio, as this will be an invaluable resource throughout this course (and in your work).

Now back to our data - you might have noticed that the columns contain a variety of income *levels*..so you might consider *income* to be a *variable* in its own right. Furthermore, an individual observation will be a combination of a `religion` (another *variable*) crossed with a specific income level.

What we want to accomplish here is to *gather* our data so that we can separate these *variables* out into `religion`, a new *key* called `income` (which is the variable that describes the income level), and then a new column where we put in the number of times we observe this specific religion and income level, which we can just refer to as `value`. 

To do this in one line, let's use the [`tidyr::gather()`](http://www.rdocumentation.org/packages/tidyr/functions/gather) function from the `tidyr` package. `gather()` takes four arguments:

- the dataset of interest 
- the name of what will be the new *key* column 
- the name of what will be the new *value* column
- the column indices where the *keys* and *values* are currently stored in the dataset

Essentially, what we end up doing is taking our dataset from a *wide* format to a *long* format. This *long* format is much friendlier with downstream applications such as `ggplot2` for constructing data visualizations (we will cover `ggplot2` in chapter 4).  

Check out the sample code which uses the built-in `iris` dataset to see `tidyr::gather()` in action.  


*** =instructions
- Read and run the first sample code chunk to inspect how `tidyr::gather()` works
- Similar to the `iris` dataset in the sample code, refactor the `pew` dataset such that the data is tidy.
- Name the *key* column `income`, and name the *value* column as just `value`
- Note that the values in the `pew` dataset are stored in columns 2:11.

*** =hint
See the sample code and your data wrangling cheat sheet for an example of how to use [`tidyr::gather()`](http://www.rdocumentation.org/packages/tidyr/functions/gather). You'll need to supply four arguments to `tidyr::gather()` - the dataset, the name of the new key column (`income`), the name of the new value column, and the column indices where values are stored.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/pew.RData"))
```

*** =sample_code
```{r}
library(tidyr)

# Example of gathering variables with the iris dataset
# The arguments are, from left to right:
# - the dataset, `iris`
# - the name for a new *key* column, let's call it "measurement" 
#     this will be a new column with values 'Sepal.Length', 'Petal.Width', etc.
# - the name for a new *value* column, where the "value"
#     this will be the column which contains the actual measurements (2.2, 3.0, etc)
# - the *column indices* values are located in the original dataset 
#     in `iris`, the values are spread across columns 1 through 4
#     you'll note in the output the column names become the *key* column (measurement)
#     and that the *value* column takes on the actual measurements from the matrix
iris_df <- tidyr::gather(iris, measurement, value, 1:4)
iris_df[1:3, ]

## Your task:
## Take the pew data set and use tidyr::gather to organize
## the data by the *key* `income`, and a *value* column named `value`
## Save the tidied dataset as pew_df
## Note that pew has its values stored in columns 2:11
pew_df


```

*** =solution
```{r}
library(tidyr)

pew_df <- tidyr::gather(pew, income, value, 2:11)
```

*** =sct
```{r}
test_object("pew_df",
            undefined_msg = "Make sure to define `pew_df`!",
            incorrect_msg = "Have you munged `pew`?")
success_msg("Good job! Head over to the next exercise!")

```

--- type:NormalExercise lang:r xp:100 skills:1,4 key:31d35975ad
## Joining Multiple Data Sets

Many times, we will encounter datasets that exist in multiple tables. While in this format we may be able to achieve a minimal representation of the data with less duplicated columns, we also make it much more difficult to actually work with the data!

Luckily for us, the package `dplyr` has a family of *join* functions that make the task of gluing data back together easier than ever. A swiss army knife of joins are available for your use - see your [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf). Check out the sample code at right to see a join in action.


*** =instructions
Use a full join to combine `dat2` with `dat3`, and save the result as `dat4`.

*** =hint
You need to identify what column you are joining *by*, and in this case, note that `dat2` and `dat3` don't have the same column name for what you want to join by (the letters of the alphabet). Check out the sample code and linked help files to see what to do in this case, using [`dplyr::full_join()`](http://www.rdocumentation.org/packages/dplyr/functions/full_join) instead of the [`dplyr::inner_join()`](http://www.rdocumentation.org/packages/dplyr/functions/inner_join) function.

*** =sample_code
```{r}
library(dplyr)

## Toy data sets
dat1 <- data.frame(id = c("a", "b", "d"), reads = c(1, 2, 4))
dat2 <- data.frame(id = c("b", "e", "f", "z"), counts = c(3, 9, 10, 25))
dat3 <- data.frame(ALPHABET = c("d", "y", "z"), times = c(7, 7, 7))

## What to do if key columns don't have matching names:
dplyr::inner_join(dat1, dat3, by = c("id" = "ALPHABET"))

## Your task:
## Combine dat2 with dat3 via a full join
```

*** =solution
```{r}
library(dplyr)

## Toy data sets
dat1 <- data.frame(id = c("a", "b", "d"), reads = c(1, 2, 4))
dat2 <- data.frame(id = c("b", "e", "f", "z"), counts = c(3, 9, 10, 25))
dat3 <- data.frame(ALPHABET = c("d", "y", "z"), times = c(7, 7, 7))

## What to do if key columns don't have matching names:
dplyr::inner_join(dat1, dat3, by = c("id" = "ALPHABET"))

## Your task:
## Combine dat2 with dat3 via a full join and save as dat4
dat4 <- dplyr::full_join(dat2, dat3, by = c("id" = "ALPHABET"))
```

*** =sct
```{r}
test_object("dat4",
            undefined_msg = "Make sure to define `dat4`!",
            incorrect_msg = "Have you munged `dat4`?")
success_msg("Good job! Head over to the next exercise!")
```

--- type:NormalExercise lang:r xp:100 skills:1,4 key:91e31433af
## Filtering Down To The Most Interesting Data 

Sometimes, we only want to look at the most interesting data in the world. To do that, we need to know how to *filter* our dataset. Thankfully, we have a handy [`dplyr::filter()`](http://www.rdocumentation.org/packages/dplyr/functions/filter) function that does just that! 

The way we filter data is we first must define what we are interested in. For example, looking at the `iris` dataset (take a gander at it via the sample code at right), we might want:

- A specific species of flower
- Flowers whose sepals are longer than x
- Flowers whose petals are wider than y
- Any combination of the above!

To do this, we use *conditionals*, such as less-than (`<`), greater-than (`>`), equals (`==`), or not-equals (`!=`) which result in *logical* (boolean) values. Then, our [`dplyr::filter()`](http://www.rdocumentation.org/packages/dplyr/functions/filter) function will only keep what would evaluate to `TRUE` in our comparison. 

Multiple filters can be done simultaneously by specifying multiple conditionals to [`dplyr::filter()`](http://www.rdocumentation.org/packages/dplyr/functions/filter).


*** =instructions
Filter the `iris` dataset such that it only contains flowers whose petal width is less than `0.2` and sepal width is more than `3.0`. Save the the result as `iris_filtered`

*** =hint
Remember, [`dplyr::filter()`](http://www.rdocumentation.org/packages/dplyr/functions/filter) works such that it is `dplyr::filter(data, conditional1, conditional2, ...)`. Your conditional should be formatted as *Column Name* < *Value* (note no quotes on column name). Inspect the `iris` data to make sure you're using the correct column.

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}
library(dplyr)

## Take a look again at the iris dataset, note the column names
head(iris)

## Example of filtering using an equality - we only want value == 10
df <- data.frame(letters = c("a", "b", "c", "d"),
                 value = c(1, 2, 10, 5))
dplyr::filter(df, value == 10)  

## Example of using two filters - we can separate multiple filters
## using just a comma `,` - note that multiple filters are combined
## as an AND (`&`) statement
dplyr::filter(df, value > 3, value < 7) 


## Your task:
## Filter iris such that it only contains flowers whose petal width 
## is less than 0.2  and whose sepal width is more than 3.0

```

*** =solution
```{r}
library(dplyr)

## Take a look again at the iris dataset, note the column names
head(iris)

## Example of filtering using an equality - we only want value == 10
df <- data.frame(letters = c("a", "b", "c", "d"),
                 value = c(1, 2, 10, 5))
dplyr::filter(df, value == 10)  # Note we kept the 3rd row!


## Your task:
## Filter iris such that it only contains flowers whose petal width 
## is less than 0.2 
iris_filtered <- dplyr::filter(iris, Petal.Width < 0.2, Sepal.Width > 3.0)
```

*** =sct
```{r}
test_object("iris_filtered",
            undefined_msg = "Make sure to define `iris_filtered`!",
            incorrect_msg = "Have you munged `iris_filtered`?")
success_msg("Good job! That was the last exercise in the chapter. Now that you understand the basics of manipulating data, you will next learn how to visualize data!")
```
