--- 
title_meta  : Chapter 1
title       : Intro to basics
description : "In this chapter, based on DataCamp's Intro To R course, you will take your first steps with R. You will learn how to use the console as a calculator and how to assign variables. You will also get to know the basic data types in R. Let's get started!"

--- type:NormalExercise xp:100 skills:1 key:7b33557562
## How it works

In the editor on the right you should type R code to solve the exercises. When you hit the 'Submit Answer' button, every line of code is interpreted and executed by R and you get a message whether or not your code was correct. The output of your R code is shown in the console in the lower right corner.

R makes use of the `#` sign to add comments, so that you and others can understand what the R code is about. Just like Twitter! Comments are not run as R code, so they will not influence your result. For example, _Calculate 3 + 4_ in the editor on the right is a comment.

You can also execute R commands straight in the console. This is a good way to experiment with R code, as your submission is not checked for correctness.

<!-- This chapter is based on Chapter 1 of DataCamp's [Intro To R course](https://www.datacamp.com/courses/free-introduction-to-r) -->

*** =instructions
- In the editor on the right there is already some sample code. Can you see which lines are actual R code and which are comments?
- Add a line of code that calculates the sum of 6 and 12, and hit the 'Submit Answer' button.

*** =hint
Just add a line of R code that calculates the sum of 6 and 12, just like the example in the sample code!

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Calculate 3 + 4
3 + 4

# Calculate 6 + 12

```

*** =solution
```{r}
# Calculate 3 + 4
3 + 4

# Calculate 6 + 12
6 + 12
```

*** =sct
```{r}
test_output_contains("18", incorrect_msg = "Make sure to add `6 + 12` on a new line. Do not start the line with a `#`, otherwise your R code is not executed!")
success_msg("Awesome! See how the console shows the result of the R code you submitted? Now that you're familiar with the interface, let's get down to R business!")
```

--- type:NormalExercise xp:100 skills:1 key:15143ba0d7
## Arithmetic with R

In its most basic form, R can be used as a simple calculator. Consider the following arithmetic operators:

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `^`
- Modulo: `%%`

The last two might need some explaining:
- The `^` operator raises the number to its left to the power of the number to its right: for example `3^2` is 9.
- The modulo returns the remainder of the division of the number to the left by the number on its right, for example 5 modulo 3 or `5 %% 3` is 2.

With this knowledge, follow the instructions below to complete the exercise.

*** =instructions
- Type `2^5` in the editor to calculate 2 to the power 5.
- Type `28 %% 6` to calculate 28 modulo 6.
- Click 'Submit Answer' and have a look at the R output in the console.
- Note how the `#` symbol is used to add comments on the R code.

*** =hint
Another example of the modulo operator: `9 %% 2` equals `1`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# An addition
5 + 5 

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Exponentiation


# Modulo

```

*** =solution
```{r}
# An addition
5 + 5

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Exponentiation
2 ^ 5

# Modulo
28 %% 6
```

*** =sct
```{r}
msg = "Do not remove the other arithmetic examples!"
test_output_contains("2^5", incorrect_msg = "The exponentiation example is not correct. Write `2 ^ 5` on a new line.")
test_output_contains("28 %% 6", incorrect_msg = "There seems to be an issue with the modulo example. Write `28 %% 6` on a new line.")
success_msg("Great! Head over to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 key:2b39bce1c1
## Variable assignment 

You already know the basic concept of a **variable**.

In R, variable assignment is performed with the `<-` operator.

You can assign a value 4 to a variable `my_var` with the command

```
my_var <- 4
```

*** =instructions
Over to you: complete the code in the editor such that it assigns the value **42** to the variable `x` in the editor. Click 'Submit Answer'. Notice that when you ask R to print `x`, the value 42 appears.

*** =hint
Look at how the value 4 was assigned to `my_variable` in the exercise's assignment. Do the exact same thing in the editor, but now assign 42 to the variable `x`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Assign the value 42 to x
x <- 

# Print out the value of the variable x
x
```

*** =solution
```{r}
# Assign the value 42 to x
x <- 42

# Print out the value of the variable x
x
```

*** =sct
```{r}
test_object("x", undefined_msg = "Make sure to define a variable `x`.",
            incorrect_msg = "Make sure that you assign the correct value to `x`.") 
success_msg("Good job! Have you noticed that R does not print the value of a variable to the console when you did the assignment? `x <- 42` did not generate any output, because R assumes that you will be needing this variable in the future. Otherwise you wouldn't have stored the value in a variable in the first place, right? Proceed to the next exercise!")
```

--- type:NormalExercise xp:100 skills:1 key:73b70d4e32
## Basic data types in R

R works with numerous data types. Some of the most basic types to get started are:

- Decimals values like `4.5` are called **numerics**.
- Boolean values (`TRUE` or `FALSE`) are called **logical**.
- Text (or string) values are called **characters**.

Note how the quotation marks on the right indicate that "some text" is a character string.

You can check the data type of a variable with the `class()` function, as the code on the right shows.

*** =instructions
Complete the code in the editor and also print out the classes of `my_character` and `my_logical`. 

*** =hint
The code that prints the data type of `my_numeric` is already included; do a similar things for `my_character` and `my_logical`. 

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Declare variables of different types
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE 

# Check class of my_numeric
class(my_numeric)

# Check class of my_character


# Check class of my_logical

```

*** =solution
```{r}
# Declare variables of different types:
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE

# Check class of my_numeric
class(my_numeric)

# Check class of my_character
class(my_character)

# Check class of my_logical
class(my_logical)
```

*** =sct
```{r}
msg <- "Do not change the declaration of the variables!"
lapply(c("my_numeric", "my_character", "my_logical"), test_object, undefined_msg = msg, incorrect_msg = msg)
patt <- "Have you included `class(%1$s)` to print out the data type of `%1$s`?"
test_output_contains("class(my_numeric)",
                     incorrect_msg = "Do not remove the code that prints out the type of `my_numeric`.")
test_output_contains("class(my_character)",
                     incorrect_msg = sprintf(patt, "my_character"))
test_output_contains("class(my_logical)",
                     incorrect_msg = sprintf(patt, "my_logical"))
success_msg("Congratulations! This was the last exercise for this chapter. Head over to the next chapter to get immersed in the world of vectors!")
```

--- type:NormalExercise xp:100 skills:1
## What's a data frame?

A **data frame** is a commonly used data type in R that stores data of different types. You will often find yourself working with data sets that contain different data types instead of only one.

A data frame has the variables of a data set as columns and the observations as rows.

*** =instructions
Click 'Submit Answer'. The data from the built-in example data frame [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) will be printed to the console.

*** =hint
Just click 'Submit Answer' and witness the magic!

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Print out built-in R data frame
mtcars 
```

*** =solution
```{r}
# Print out built-in R data frame
mtcars 
```

*** =sct
```{r}
test_output_contains("mtcars", incorrect_msg = "Do not change anything about the code, Make sure that you output `mtcars`.")
success_msg("Great! Continue to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 
## Quick, have a look at your data set

Wow, that is a lot of cars! 

Working with large data sets is not uncommon in data analysis. When you work with (extremely) large data sets and data frames, your first task as a data analyst is to develop a clear understanding of its structure and main elements. Therefore, it is often useful to show only a small part of the entire data set. 

So how to do this in R? Well, the function [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) enables you to show the first observations of a data frame. Similarly, the function [`tail()`](http://www.rdocumentation.org/packages/utils/functions/head) prints out the last observations in your data set.

Both [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) and [`tail()`](http://www.rdocumentation.org/packages/utils/functions/head) print a top line called the 'header', which contains the names of the different variables in your data set.

*** =instructions
Call [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) on the [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) data set to have a look at the header and the first observations.

*** =hint
`head(mtcars)` will show the first observations of the `mtcars` data frame.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Call head() on mtcars

```

*** =solution
```{r}
# Call head() on mtcars
head(mtcars)
```

*** =sct
```{r}
test_function("head", "x", incorrect_msg = "Have you correctly passed `mtcars` to the `head()` function?")
test_output_contains("head(mtcars)", incorrect_msg = "Simply print out the result of the `head()` call, no need to assign it to a new variable.")
success_msg("Wonderful! So, what do we have in this data set? For example, `hp` for example represents the car's horsepower; the Datsun has the lowest horse power of the 6 cars that are displayed. For a full overview of the variables' meaning, type `?mtcars` in the console and read the help page. Continue to the next exercise!");
```


--- type:NormalExercise xp:100 skills:1 
## Have a look at the structure

Another method that is often used to get a rapid overview of your data is the function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str). The function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) shows you the structure of your data set. For a data frame it tells you:

- The total number of observations (e.g. 32 car types)
- The total number of variables (e.g. 11 car features)
- A full list of the variables names (e.g. `mpg`, `cyl` ... )
- The data type of each variable (e.g. `num`)
- The first observations

Applying the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function will often be the first thing that you do when receiving a new data set or data frame. It is a great way to get more insight in your data set before diving into the real analysis.

*** =instructions
Investigate the structure of [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars). Make sure that you see the same numbers, variables and data types as mentioned above.

*** =hint
Use the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function on [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars).

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Investigate the structure of mtcars
```

*** =solution
```{r}
# Investigate the structure of mtcars
str(mtcars)
```

*** =sct
```{r}
test_output_contains("str(mtcars)", incorrect_msg = "Have you correctly called `str()` on `mtcars`?")
success_msg("Nice work! Can you find all the information that is listed in the exercise's assignment? Continue to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1
## Creating a data frame 

Since using built-in data sets is not even half the fun of creating your own data sets, the rest of this chapter is based on your personally developed data set. Put your jet pack on because it is time for some space exploration! 

As a first goal, you want to construct a data frame that describes the main characteristics of eight planets in our solar system. According to your good friend Buzz, the main features of a planet are:

- The type of planet (Terrestrial or Gas Giant).
- The planet's diameter relative to the diameter of the Earth.
- The planet's rotation across the sun relative to that of the Earth.
- If the planet has rings or not (TRUE or FALSE).

After doing some high-quality research on [Wikipedia](http://en.wikipedia.org/wiki/Planet), you feel confident enough to create the necessary vectors: `name`, `type`, `diameter`, `rotation` and `rings`; these vectors have already been coded up on the right. The first element in each of these vectors correspond to the first observation.

You construct a data frame with the [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) function. As arguments, you pass the vectors from before: they will become the different columns of your data frame. Because every column has the same length, the vectors you pass should also have the same length. But don't forget that it is possible (and likely) that they contain different types of data.

*** =instructions
Use the function [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) to construct a data frame. Pass the vectors `name`, `type`, `diameter`, `rotation` and `rings` as arguments to `data.frame()`, in this order. Call the resulting data frame `planets_df`.

*** =hint
Your `data.frame()` call starts as follows:
```
data.frame(planets, type, diameter)
```
Can you finish it?

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <-

```

*** =solution
```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <- data.frame(name, type, diameter, rotation, rings)
```

*** =sct
```{r}
msg = "Do not change anything about the definition of the vectors. Only add a `data.frame()` call to create `planets_df`."
test_object("name", undefined_msg = msg, incorrect_msg = msg)
test_object("type", undefined_msg = msg, incorrect_msg = msg)
test_object("diameter", undefined_msg = msg, incorrect_msg = msg)
test_object("rotation", undefined_msg = msg, incorrect_msg = msg)
test_object("rings", undefined_msg = msg, incorrect_msg = msg)

test_object("planets_df",
            incorrect_msg = "Have you correctly called `data.frame()` to create `planets_df`. Inside `data.frame()`, make sure to pass all vectors in the correct order: `name`, `type`, `diameter`, `rotation` and finally `rings`.")

success_msg("Great job! Continue to the next exercise. The logical next step, as you know by now, is inspecting the data frame you just created. Head over to the next exercise.");
```
