--- 
title_meta  : Chapter 4
title       : Application to Expression Data
description : "In this chapter, you will combine what you learned in the first three chapters to read in, tidy, and visualize gene expression data. Let's get started!"

--- type:NormalExercise xp:100 skills:1
## Putting it all together

Now that you've learned how to work with R, use `tidyr`/`dplyr` and `ggplot2` you will apply these skills to read in and visualize gene expression data.

To get microarray data, a handy package `GEOquery` exists on Bioconductor to read in the data. We will read expression data for study [GSE55268](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE55278). This study describes expression profiling of dendritic cells after infection with multiple strains of influenza viruses over an 8-hour time course. Read the abstract of [the associated publication](http://www.ncbi.nlm.nih.gov/pubmed/26223639) before continuing.

*** =instructions
- Hit Submit Answer to continue

*** =hint
There is no code to type, just hit Submit Answer.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
```

*** =solution
```{r}
```

*** =sct
```{r}
success_msg("Now that you understand the study, let's analyze some data!")
```

