---
title_meta  : Chapter 4
title       : Application to Expression Data
description : "In this chapter, you will combine what you learned in the first three chapters to read in, tidy, and visualize gene expression data. Let's get started!"

--- type:NormalExercise xp:25 key:2200ac0aae
## Putting it all together

Now that you've learned how to work with R, use `tidyr`/`dplyr` and `ggplot2` you will apply these skills to read in and visualize gene expression data.

We will read expression data for study [GSE55268](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE55278). This study describes expression profiling of dendritic cells after infection with multiple strains of influenza viruses over an 8-hour time course. Read the abstract of [the associated publication](http://www.ncbi.nlm.nih.gov/pubmed/26223639) before continuing.

*** =instructions
- Hit 'Submit Answer' to continue

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
```

*** =solution
```{r}
```

*** =sct
```{r}
success_msg("Now that you understand the study, let's get the data!")
```

--- type:NormalExercise xp:25 skills:1,8 key:ecc45725c8
## Reading in the data
To get expression data from the Gene Expression Omnibus (GEO), you can use a package called `GEOquery` from Bioconductor. The size of the data prohibits us from running the code inside DataCamp but it is very simple to do in a standard R environment.

*** =instructions
- Look at the code required to read expression data from GEO.
- Hit 'Submit Answer' to continue

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# library(GEOquery)
# geoList <- getGEO("GSE55278")
# eset <- geoList[[1]]
```

*** =solution
```{r}
# library(GEOquery)
# geoList <- getGEO("GSE55278")
# eset <- geoList[[1]]
```

*** =sct
```{r}
success_msg("Doesn't it look easy to read in data!?  Because the data is large, you'll be provided with a subset of the data in `eset` for the rest of the chapter.")
```

--- type:NormalExercise  xp:100 skills:1 key:483c4c5eb5
## ExpressionSet Objects

`GEOQuery` returns data in an object of the `ExpressionSet` class.  You can think of an `ExpressionSet` class as a container designed to combine multiple sources of information from a gene expression experiment into a single object. If interested, you can read more about this class [here](https://www.bioconductor.org/packages/devel/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf).

For this course, you need to know three functions to extract data from an `ExpressionSet`

- `pData`: Retreive information on experimental phenotypes (e.g. geo\_accession, Condition, Time)
- `fData`: Retrieve information on measured features (e.g. Entrez\_Gene\_ID, Symbol, Species)
- `exprs`: Retrieve expression data matrix

*** =instructions
- Create a new variable (`expr`) that contains the expression data from `eset`
- Create a new variable (`features`) that contains the feature data from `eset`
- Create a new variable (`phenotypes`) that contains the phenotype data from `eset`
- Print `eset` to get an overview of the data it contains.
- Click 'Submit Answer'.

*** =hint
Remember that if you want to assign values to a variable in R, you can make use of the assignment operator `<-`. Use the functions listed above to extract data from `eset`.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/eset.RData"))
```

*** =sample_code
```{r}
# Load Biobase package (from Bioconductor) to work with ExpressionSet Class
library(Biobase)

# Extract information out of eset into 3 new variables




# Print eset
```

*** =solution
```{r}
# Load Biobase package (from Bioconductor) to work with ExpressionSet Class
library(Biobase)

# Extract information out of eset into 3 new variables
expr <- exprs(eset)
features <- fData(eset)
phenotypes <- pData(eset)

# Print eset
print(eset)
```

*** =sct
```{r}
test_student_typed(c("print(eset)", "eset", "show(eset)"),
                    fixed = TRUE, times = 1,
                    not_typed_msg = "Have you printed the `eset` variable?")
# test_output_contains(print(eset), incorrect_msg = "Have you printed the `eset` variable?")
test_object("expr",
            undefined_msg = "Please make sure to define a variable `expr`.",
            incorrect_msg = "Have you set the variable `expr`?")
test_object("features",
            undefined_msg = "Please make sure to define a variable `features`.",
            incorrect_msg = "Have you set the variable `features`?")
test_object("phenotypes",
            undefined_msg = "Please make sure to define a variable `phenotypes`.",
            incorrect_msg = "Have you set the variable `phenotypes`?")
success_msg("Nice job! Notice in the output that printing `eset` tells you about what it contains including the number of features (genes) and samples.")
```

--- type:NormalExercise  xp:100 skills:1 key: key:d454b2036d
## Filter the Data

Now that we extracted our data from the Expression Set, we can take a quick look.


*** =instructions

- Print out the column names of `phenotypes` and `features`
- Use `dplyr::select()` to select `geo_accession`, `title`, `Treatment`, `Time`, and `Replicate` from `phenotypes` and save the result in `phenotypes_filtered`
- Use `dplyr::select()` to select `ID`, `Entrez_Gene_ID`, and `Symbol` from `features` and save the result in `features_filtered`


*** =hint
Use the `colnames()` function to look at the column names of `phenotypes` and `features`. Then use `dplyr::select()` to retain a subset of the columns.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
```

*** =sample_code
```{r}
library(dplyr)

# Print column names of phenotypes and features

# Only keep subset of columns needed for analysis
```

*** =solution
```{r}
library(dplyr)

# Print column names of phenotypes and features
colnames(phenotypes)
colnames(features)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)
```

*** =sct
```{r}
test_output_contains("colnames(phenotypes)",
                     incorrect_msg = "Did you remember to print out the column names of `phenotypes`?")
test_output_contains("colnames(features)",
                     incorrect_msg = "Did you remember to print out the column names of `features`?")
test_object("phenotypes_filtered",
            undefined_msg = "Make sure to name your new variable `phenotypes_filtered`.",
            incorrect_msg = "Have you selected the corrected columns from `phenotypes`?")
test_object("features_filtered",
            undefined_msg = "Make sure to name your new variable `features_filtered`.",
            incorrect_msg = "Have you selected the corrected columns from `features`?")
success_msg("Great! Continue to the next exercise.")
```



--- type:NormalExercise  xp:100 skills:1 key:e7416248d9
## Combine the data

Currently our data is stored in 3 variables, `expr`, `features_filtered`, and `phenotypes_filtered`. Eventually we want to combine all of these into a single tidy data frame.

To begin, we can combine the information on the features with the expression data. Each row of `features_filtered` gives additional information on the gene expression probe while each row of `expr` gives the expression levels of each probe in different samples.

We can combine these together in a data frame using the [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) function.

*** =instructions

- Pass `features_filtered` and `expr` to the [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) function and assign this new data frame to a variable named `expr_df`.
- Click 'Submit Answer'.

*Note: We are ignoring the `factor` data type by setting the global option `stringsAsFactors` to `FALSE` in the R workspace.*

*** =hint
The [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) function can take one of more vectors, matrices or data.frames and will combining by binding columns together. A vector will be treated like a 1-D matrix.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
options(stringsAsFactors = FALSE)
```

*** =sample_code
```{r}
library(dplyr)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)

# Combine features and expr into a data frame named expr_df
```

*** =solution
```{r}
library(dplyr)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)

# Combine features and expr into a data frame named expr_df
expr_df <- data.frame(features_filtered, expr)
```

*** =sct
```{r}
test_object("expr_df",
            undefined_msg = "Please make sure to define a variable `expr_df`.",
            incorrect_msg = "Have you set the variable `expr_df`?")
success_msg("Great! Continue to the next exercise to continue tidying this expression data.")
```

--- type:NormalExercise xp:100 skills:1,4 key:f53f507908
## Tidy the data

Next you'll need to transform your data frame into a tidy format using [`tidyr::gather()`](http://www.rdocumentation.org/packages/tidyr/functions/gather) so that there is only one observation (i.e. one gene measured in one sample) per row.

*** =instructions

- Use [`tidyr::gather()`](http://www.rdocumentation.org/packages/tidyr/functions/gather) on `expr_df` specifying the key as `geo_accession` and the value as `expression` and save this in a variable called `dat`.
  - You will have to specify in `tidyr::gather()` which columns should be gathered (expression samples start with "GSM")
- Look at the structure of `dat` using the `str()` function

*** =hint
The [`tidyr::gather()`](http://www.rdocumentation.org/packages/tidyr/functions/gather) function requires a key, value, and specification of which columns to gather.

You can specify all of the samples named by geo\_accession using `starts_with("GSM")`

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
options(stringsAsFactors = FALSE)
```

*** =sample_code
```{r}
library(tidyr)
library(dplyr)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)

# Combine features_filtered and expr into a data frame named expr_df
expr_df <- data.frame(features_filtered, expr)

# Gather expression columns of expr_df and print structure of dat
```

*** =solution
```{r}
library(tidyr)
library(dplyr)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)

# Combine features_filtered and expr into a data frame named expr_df
expr_df <- data.frame(features_filtered, expr)

# Gather expression columns of expr_df and print structure of dat
dat <- tidyr::gather(data = expr_df, key = geo_accession,
              value = expression, starts_with("GSM"))
str(dat)
```

*** =sct
```{r}
## Testing the object here is too resource intensive and results in session disconnects
## test_object("dat",
##             undefined_msg = "Please make sure to define a variable `dat`.",
##             incorrect_msg = "Looks like `dat` is not set properly. Check your
## `tidyr::gather()` call")
## test_output_contains(str(dat),
##                      incorrect_msg = "Have you correctly gathered the data into `dat` and printed the structure of `dat`?")
test_object("dat", eval = FALSE)
success_msg("Your dataset should contain 5 variables (`ID`, `Entrez_Gene_ID`, `Symbol`, `geo_accession`, `expression`) and 46,560 observations. If you did this correctly, continue to the next exercise. If not, look at the hint and try again!")
```

--- type:NormalExercise xp:100 skills:1,4 key:ebeb4cda30
## Join the data

Finally, you want to add phenotype data in `phenotypes_filtered` to `dat` so you have all the information describing the geo_accession (i.e. sample) in each row.

[`full_join()`](http://www.rdocumentation.org/packages/dplyr/functions/full_join) can be used to join two data frames and retain all values, and all rows.


*** =instructions

- Use [`full_join()`](http://www.rdocumentation.org/packages/dplyr/functions/full_join) to join `dat` and `phenotypes_filtered` by the "geo_accession" column and save the result in a new variabled called `data`

*** =hint

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
options(stringsAsFactors = FALSE)
```

*** =sample_code
```{r}
library(tidyr)
library(dplyr)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)

# Combine features_filtered and expr into a data frame named expr_df
expr_df <- data.frame(features_filtered, expr)

# Gather expression columns of expr_df and print structure of dat
dat <- tidyr::gather(data = expr_df, key = geo_accession,
              value = expression, starts_with("GSM"))

# Join dat and phenotypes_filtered
```

*** =solution
```{r}
library(tidyr)
library(dplyr)

# Only keep subset of columns needed for analysis
phenotypes_filtered <- dplyr::select(phenotypes, geo_accession, title,
                                     Treatment, Time, Replicate)
features_filtered <- dplyr::select(features, ID, Entrez_Gene_ID, Symbol)

# Combine features_filtered and expr into a data frame named expr_df
expr_df <- data.frame(features_filtered, expr)

# Gather expression columns of expr_df and print structure of dat
dat <- tidyr::gather(data = expr_df, key = geo_accession,
              value = expression, starts_with("GSM"))

# Join dat and phenotypes_filtered
data <- full_join(dat, phenotypes_filtered, by = "geo_accession")
```

*** =sct
```{r}
test_object("data",
            undefined_msg = "Please make sure to define a variable `data`.",
            incorrect_msg = "Looks like `data` is not set properly. Check your `full_join()` call")
success_msg("Great! Now that your data is tidy, it's time for the fun part - exploring and visualizing the data!")
```

--- type:NormalExercise xp:200 skills:1 key:9c603bbce2
## Plotting Gene Expression

You will start visually exploring by plotting the expression of one gene over time during influenza infection.


*** =instructions

- Plot the `expression` of the gene *MX1* over `Time` during infection with the *Brevig* strain.

  - First [`dplyr::filter()`](http://www.rdocumentation.org/packages/dplyr/functions/filter) `data` to observations where the `Symbol` is "MX1" and the `Treatment` is "Brevig" and save this in a new variable.
  - Then call `ggplot()` on this new variable and add `x` and `y` aesthetics using `aes()`.
  - Finally, add a `geom_point()` layer to the `ggplot()` call.

*** =hint

Remember to use `+` to add layers to the plotting call.

Review Chapter 3 if you forget what a `ggplot()` call should look like.

*** =pre_exercise_code
```{r}

load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
options(stringsAsFactors = FALSE)
```

*** =sample_code
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Tidy data
# Note we are doing the same steps here using piping (%>%)
data <- data.frame(features, expr) %>%
  gather(key = geo_accession, value = expression, starts_with("GSM")) %>%
  full_join(phenotypes, by = "geo_accession") %>%
  select(geo_accession, title, Treatment, Time, Replicate,
         ID, Entrez_Gene_ID, Symbol, expression)

# Filter the data


# Plot
```

*** =solution
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Tidy data
# Note we are doing the same steps here using piping (%>%)
data <- data.frame(features, expr) %>%
  gather(key = geo_accession, value = expression, starts_with("GSM")) %>%
  full_join(phenotypes, by = "geo_accession") %>%
  select(geo_accession, title, Treatment, Time, Replicate,
         ID, Entrez_Gene_ID, Symbol, expression)

# Filter the data
plotData <- dplyr::filter(data, Symbol == "MX1", Treatment == "Brevig")

# Plot
ggplot(data = plotData, aes(x = Time, y = expression)) +
  geom_point() +
  theme_bw()

############################################################
# # In comparison, the same plot can be made without tidying
# # the data first but it is less readable and more code.
# cond <- phenotypes[, "Treatment"] == "Brevig"
# exp <- expr[features_filtered[,"Symbol"] == "MX1", cond]
# times <- phenotypes[cond, "Time"]
# plot(x = times, y = exp, pch = 19)
```

*** =sct
```{r}
test_ggplot(check_data = TRUE, data_fail_msg = "Make sure to provide the filtered data",
            check_aes = TRUE, aes_fail_msg = "Did you specify the `x` and `y` aesthetics in `aes()` to plot `Time` versus `expression`?",
            check_geom = TRUE, geom_fail_msg = "Did you add a `geom_point()` layer to create a scatter plot?")
success_msg("Fantastic!  You just created a scatter plot of expression over time! You can see that the expression of *MX1* changes dramatically over the time course. That's because *MX1* is an antiviral response gene that gets activated when a cell is infected. Let's spice it up a bit...")
```

--- type:NormalExercise xp:200 skills:1 key:f548bed685
## Plotting Gene Expression (2)

You just plotted the gene expression profile of *MX1* during an influenza infection! Did you notice there were three points at each time point? These correspond to 3 biological replicates. To see the trajectory of each replicate, you can connect each replicate over time with a line.

The great thing about `ggplot2` is that tweaking the visualization is as easy as adding more layers to your plot. To see this, add a line to your scatter plot connecting the replicates over time.


*** =instructions

- Add a `geom_line()` layer to the existing plot and specify the `group` aesthetic for `geom_line()` to be `Replicate`.


*** =hint

Aesthetics can be specified within each layer. Your code should look something like this: `geom_line(aes(...))`

Remember you can add a layer using `+`

*** =pre_exercise_code
```{r}

load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
options(stringsAsFactors = FALSE)
```

*** =sample_code
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Tidy data
# Note we are doing the same steps here using piping (%>%)
data <- data.frame(features, expr) %>%
  gather(key = geo_accession, value = expression, starts_with("GSM")) %>%
  full_join(phenotypes, by = "geo_accession") %>%
  select(geo_accession, title, Treatment, Time, Replicate,
         ID, Entrez_Gene_ID, Symbol, expression)

# Filter the data
plotData <- dplyr::filter(data, Symbol == "MX1", Treatment == "Brevig")

# Add a line to this plot
ggplot(data = plotData, aes(x = Time, y = expression)) +
  geom_point()

```

*** =solution
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Tidy data
# Note we are doing the same steps here using piping (%>%)
data <- data.frame(features, expr) %>%
  gather(key = geo_accession, value = expression, starts_with("GSM")) %>%
  full_join(phenotypes, by = "geo_accession") %>%
  select(geo_accession, title, Treatment, Time, Replicate,
         ID, Entrez_Gene_ID, Symbol, expression)

# Filter the data
plotData <- dplyr::filter(data, Symbol == "MX1", Treatment == "Brevig")

# Add a line to this plot
ggplot(data = plotData, aes(x = Time, y = expression)) +
  geom_point() +
  geom_line(aes(group = Replicate)) +
  theme_bw()

############################################################
# # In comparison, the same plot can be made without tidying
# # the data first but it is less readable and more code.
# cond <- phenotypes[, "Treatment"] == "Brevig"
# exp <- expr[features[,"Symbol"] == "MX1", cond]
# times <- phenotypes[cond, "Time"]
# matExp <- matrix(exp, ncol = 3, byrow = TRUE)
# matTimes <- matrix(times, ncol = 3, byrow = TRUE)
# reps <- phenotypes[cond, "Replicate"]
# matplot(x = matTimes, y = matExp, type = 'b',
#         lty = 1, pch = 19, col = "black")
```

*** =sct
```{r}
test_ggplot(check_data = TRUE, data_fail_msg = "Make sure to provide the filtered data",
            check_aes = TRUE, aes_fail_msg = "Did you specify the `x` and `y` aesthetics in `aes()` to plot `Time` versus `expression`?",
            check_geom = TRUE, geom_fail_msg = "Did you keep the `geom_point()` layer and add a new `geom_line()` layer with a group aesthetic?")
success_msg("Lookin' good! Ready to make a publication-quality visualization?")
```

--- type:NormalExercise xp:200 skills:1 key:6a67c9447e
## Plotting Gene Expression (3)

This experiment was designed to compare different strains of the flu virus but so far you have only been plotting a single strain (*Brevig*).

Make one plot that compares the expression levels of *MX1* for all values of `Treatment`.

*** =instructions

Create a plot similar to the example plot except:

- Filter `data` only by `Symbol` to keep all values of `Treatment`
- Specify the `color` aesthetic in the `ggplot()` call to depend on `Treatment`
- Modify the `group` aesthetic in `geom_line()` to depend on both `Replicate` and `Treatment` by using `interaction(Replicate, Treatment)`.

*** =hint

Aesthetics can be specified within each layer. Your call to `geom_line()` should look something like this: `geom_line(aes(group = ...))`

Remember you can add a layer using `+`

*** =pre_exercise_code
```{r}

load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1841/datasets/extractedEset.RData"))
options(stringsAsFactors = FALSE)
```

*** =sample_code
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Tidy data
# Note we are doing the same steps here using piping (%>%)
data <- data.frame(features, expr) %>%
  gather(key = geo_accession, value = expression, starts_with("GSM")) %>%
  full_join(phenotypes, by = "geo_accession") %>%
  select(geo_accession, title, Treatment, Time, Replicate,
         ID, Entrez_Gene_ID, Symbol, expression)

# Filter the data
plotData <- dplyr::filter(data, Symbol == "MX1", Treatment == "Brevig")

# Example plot
ggplot(data = plotData, aes(x = Time, y = expression)) +
  geom_point() +
  geom_line(aes(group = Replicate)) +
  theme_bw()

############################################################
# Filter the data


# New plot
```

*** =solution
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Tidy data
# Note we are doing the same steps here using piping (%>%)
data <- data.frame(features, expr) %>%
  gather(key = geo_accession, value = expression, starts_with("GSM")) %>%
  full_join(phenotypes, by = "geo_accession") %>%
  select(geo_accession, title, Treatment, Time, Replicate,
         ID, Entrez_Gene_ID, Symbol, expression)

# Filter the data
plotData <- dplyr::filter(data, Symbol == "MX1", Treatment == "Brevig")

# Example plot
ggplot(data = plotData, aes(x = Time, y = expression)) +
  geom_point() +
  geom_line(aes(group = Replicate)) +
  theme_bw()

############################################################
# Filter the data
plotData <- dplyr::filter(data, Symbol == "MX1")

# New plot
ggplot(data = plotData, aes(x = Time, y = expression, color = Treatment)) +
  geom_point() +
  geom_line(aes(group = interaction(Replicate, Treatment))) +
  theme_bw()
```

*** =sct
```{r}
test_ggplot(index = 2,
            check_data = TRUE, data_fail_msg = "Make sure to provide the new filtered data",
            check_aes = TRUE, aes_fail_msg = "Did you specify the `x`, `y`, and `color` aesthetics as well as the `group` for `geom_line()`?",
            check_geom = TRUE, geom_fail_msg = "Did you add both a `geom_point()` layer and add a `geom_line()` layer?")
success_msg("Lookin' good! You can use the buttons on the plot to go back and forth between multiple plots. How does the `Brevig` strain compare to the other strains (`Cal`, `NC`, `Tx`)? `AlaF` is a mock infection while `Control` is no infection. Does this make sense from the plot? This was the last exercise in the CBB Jamboree Intro to R. You now have a solid basis for downloading, manipulating, transforming, and visualizing data in R!")
```
