--- 
title_meta  : Chapter 1
title       : Intro to basics
description : "In this chapter, based on DataCamp's Intro To R course, you will take your first steps with R. You will learn how to use the console as a calculator and how to assign variables. You will also get to know the basic data types in R. Let's get started!"

--- type:NormalExercise xp:100 skills:1 key:7b33557562
## How it works

In the editor on the right you should type R code to solve the exercises. When you hit the 'Submit Answer' button, every line of code is interpreted and executed by R and you get a message whether or not your code was correct. The output of your R code is shown in the console in the lower right corner.

R makes use of the `#` sign to add comments, so that you and others can understand what the R code is about. Just like Twitter! Comments are not run as R code, so they will not influence your result. For example, _An addition_ in the editor on the right is a comment.

You can also execute R commands straight in the console. This is a good way to experiment with R code, as your submission is not checked for correctness.

Consider the following arithmetic operators:

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `^`
- Modulo: `%%` (returns the remainder of the division of the number to the left by the number on its right)

For help on any built-in functions or operators, you can type `help("func_name")` in the console. For example, help("+").

*** =instructions
- In the editor on the right there is already some sample code. Can you see which lines are actual R code and which are comments?
- Add a line of code that calculates 2 raised to the power 5 and hit the 'Submit Answer' button.
*** =hint
Just add a line of R code that calculates the sum of 6 and 12, just like the example in the sample code!

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# An addition
5 + 5 

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Modulo
28 %% 6

# Exponentiation
```

*** =solution
```{r}
# An addition
5 + 5 

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Modulo
28 %% 6

# Exponentiation
2 ^ 5
```

*** =sct
```{r}
test_output_contains("32", incorrect_msg = "Make sure to raise 2 to the power 5. Do not start the line with a `#`, otherwise your R code is not executed!")
success_msg("Awesome! See how the console shows the result of the R code you submitted? Now that you're familiar with the interface, let's get down to R business!")
```

--- type:NormalExercise xp:100 skills:1 key:2b39bce1c1
## Variable assignment 

You already know the basic concept of a **variable**.

In R, variable assignment is performed with the `<-` operator.

You can assign a value 4 to a variable `my_var` with the command

```
my_var <- 4
```

*** =instructions
Over to you: complete the code in the editor such that it assigns the value **42** to the variable `x` in the editor. Click 'Submit Answer'. Notice that when you ask R to print `x`, the value 42 appears.

*** =hint
Look at how the value 4 was assigned to `my_var` in the exercise's assignment. Do the exact same thing in the editor, but now assign 42 to the variable `x`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Assign the value 42 to x
x <- ___

# Print out the value of the variable x
x
```

*** =solution
```{r}
# Assign the value 42 to x
x <- 42

# Print out the value of the variable x
x
```

*** =sct
```{r}
test_object("x", undefined_msg = "Make sure to define a variable `x`.",
            incorrect_msg = "Make sure that you assign the correct value to `x`.") 
success_msg("Good job! Have you noticed that R does not print the value of a variable to the console when you did the assignment? `x <- 42` did not generate any output, because R assumes that you will be needing this variable in the future. Otherwise you wouldn't have stored the value in a variable in the first place, right? Proceed to the next exercise!")
```

--- type:NormalExercise xp:100 skills:1 key:73b70d4e32
## Basic data types in R

R works with numerous data types. Some of the most basic types to get started are:

- Decimals values like `4.5` are called **numerics**.
- Boolean values (`TRUE` or `FALSE`) are called **logical**.
- Text (or string) values are called **characters**.

Note how the quotation marks on the right indicate that "some text" is a character string.

You can check the data type of a variable with the `class()` function, as the code on the right shows.

*** =instructions
Complete the code in the editor and also print out the classes of `my_character` and `my_logical`. 

*** =hint
The code that prints the data type of `my_numeric` is already included; do a similar things for `my_character` and `my_logical`. 

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Declare variables of different types
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE 

# Check class of my_numeric
class(my_numeric)

# Check class of my_character


# Check class of my_logical

```

*** =solution
```{r}
# Declare variables of different types:
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE

# Check class of my_numeric
class(my_numeric)

# Check class of my_character
class(my_character)

# Check class of my_logical
class(my_logical)
```

*** =sct
```{r}
msg <- "Do not change the declaration of the variables!"
lapply(c("my_numeric", "my_character", "my_logical"), test_object, undefined_msg = msg, incorrect_msg = msg)
patt <- "Have you included `class(%1$s)` to print out the data type of `%1$s`?"
test_output_contains("class(my_numeric)",
                     incorrect_msg = "Do not remove the code that prints out the type of `my_numeric`.")
test_output_contains("class(my_character)",
                     incorrect_msg = sprintf(patt, "my_character"))
test_output_contains("class(my_logical)",
                     incorrect_msg = sprintf(patt, "my_logical"))
success_msg("Nice work! Now that you know the basic data types, continue on to learn about data frames - one of the most useful data types in R!")
```

--- type:NormalExercise xp:100 skills:1 key:a79a6030f4
## What's a data frame?

A **data frame** is a commonly used data type in R that can store data of multiple types. You will often find yourself working with data sets that contain different data types instead of only one. A data frame can be created with the [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) function.

In a data frame, variables are stored in columns and observations are stored in rows. A data frame can also hold data of all the same type as in the built-in [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) data set.

*** =instructions
- Click 'Submit Answer'.
  - A toy data frame containing numeric and character values is created using two vectors (themselves created with the combine ([`c()`](http://www.rdocumentation.org/packages/base/functions/c)) function
  - Data from the built-in example data frame [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) will be printed to the console.

*** =hint
Just click 'Submit Answer' and witness the magic!

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Create a toy data frame to hold names and ages
names <- c("Doreen", "Joe", "Jamal")
ages <- c(75, 34, 23)
toy_df <- data.frame(Name = names, Age = ages)
toy_df


# Print out built-in R data frame
mtcars 
```

*** =solution
```{r}
# Create a toy data frame to hold names and ages
names <- c("Doreen", "Joe", "Jamal")
ages <- c(75, 34, 23)
toy_df <- data.frame(Name = names, Age = ages)
toy_df


# Print out built-in R data frame
mtcars 
```

*** =sct
```{r}
test_output_contains("toy_df", incorrect_msg = "Do not change anything about the code, Make sure that you output `toy_df`.")
test_output_contains("mtcars", incorrect_msg = "Do not change anything about the code, Make sure that you output `mtcars`.")
success_msg("Great! Continue to the next exercise.")
```

--- type:NormalExercise xp:100 skills:1  key:b1506e572c
## Quick, have a look at your data

Wow, that is a lot of cars! 

Working with large data sets is not uncommon in bioinformatics. When you work with (extremely) large data sets and data frames, your first task as a data analyst is to develop a clear understanding of its structure and main elements. Therefore, it is often useful to show only a small part of the entire data set. 

So how to do this in R? Well, the function [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) enables you to show the first few observations (6 by default) of a data frame. Similarly, [`tail()`](http://www.rdocumentation.org/packages/utils/functions/head) prints out the last few observations in your data frame.

*** =instructions

Call [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) on the [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) data set to have a look at the header and the first observations.



*** =hint
`head(mtcars)` will show the first observations of the `mtcars` data frame.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Call head() on mtcars

```

*** =solution
```{r}
# Call head() on mtcars
head(mtcars)
```

*** =sct
```{r}
test_function("head", "x", incorrect_msg = "Have you correctly passed `mtcars` to the `head()` function?")
test_output_contains("head(mtcars)", incorrect_msg = "Simply print out the result of the `head()` call, no need to assign it to a new variable.")

success_msg("Wonderful! So, what do we have in this data set? `hp` for example represents the car's horsepower; the Datsun has the lowest horse power of the 6 cars that are displayed. For a full overview of the variables' meaning, type `help(\"mtcars\")` in the console and read the help page. Continue to the next exercise!");
```

--- type:NormalExercise xp:100 skills:1  key:e94dbf26f4
## A second look, data structure

Another method that is often used to get a rapid overview of your data is the function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str). The function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) shows you the structure of your data set. For a data frame it tells you:

- The total number of observations (e.g. 32 car types)
- The total number of variables (e.g. 11 car features)
- A full list of the variables names (e.g. `mpg`, `cyl` ... )
- The data type of each variable (e.g. `num`)
- The first observations

Applying the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function will often be the first thing that you do when receiving a new data set or data frame. It is a great way to get more insight in your data set before diving into the real analysis.

It is also useful to summarize all of the observations in each column. You can do this in R with the [`summary()`](http://www.rdocumentation.org/packages/utils/functions/summary) function.

*** =instructions

- Investigate the structure of [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars). Make sure that you see the same numbers, variables and data types as mentioned above.

- Summarize each column of [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars).


*** =hint
Use the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function on [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars).

Use the [`summary()`](http://www.rdocumentation.org/packages/utils/functions/summary) function on [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars).


*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Investigate the structure of mtcars


# Summarize each column of mtcars
```

*** =solution
```{r}
# Investigate the structure of mtcars
str(mtcars)

# Summarize each column of mtcars
summary(mtcars)
```

*** =sct
```{r}
test_output_contains("str(mtcars)", incorrect_msg = "Have you correctly called `str()` on `mtcars`?")
test_output_contains("summary(mtcars)", incorrect_msg = "Have you correctly called `summary()` on `mtcars`?")

success_msg("Nice work! Can you find all the information in the list on this exercise's assignment? Continue to the next exercise.")
```

--- type:NormalExercise xp:100 key:86cb110648
## Slicing data frames

Data frames can be sliced (indexed) by `[row, col]` where `row` and `col` are:

- Numeric vectors giving the indices
- Logical vectors equal in length to the dimensions
- Character vectors specifying the row and column names

If either `row` or `col` is left blank, all elements will be selected. Be careful here to still include the comma. And if you only select a single column, R will convert the result to a vector (you have to specify `drop = FALSE` to keep the result as a data frame)

*** =instructions
Slice the `mtcars` data frame to extract the $10^{\textrm{th}}$ row for the `mpg` and `carb` columns.

*** =hint
Index `mtcars` with `[row, col]`

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Get row 1, all columns
mtcars[1, ]
 
# Get all rows, column 1 (result is a vector)
mtcars[,1]
# Get all rows, column 1 (result is a data frame)
mtcars[ , 1, drop = FALSE]

# Define row to be a numeric type
row <- 10
# Define col to be a character vector
col <- c("mpg", "carb")

# Slice mtcars
```

*** =solution
```{r}
# Get row 1, all columns
mtcars[1, ]
 
# Get all rows, column 1 (result is a vector)
mtcars[,1]
# Get all rows, column 1 (result is a data frame)
mtcars[ , 1, drop = FALSE]

# Define row to be a numeric type
row <- 10
# Define col to be a character vector
col <- c("mpg", "carb")

# Slice mtcars
mtcars[row, col]
```

*** =sct
```{r}
msg = "Do not change anything about the definition of the vectors. Only add a call to slice `mtcars`."
test_object("row", undefined_msg = msg, incorrect_msg = msg)
test_object("col", undefined_msg = msg, incorrect_msg = msg)

test_output_contains("mtcars[row,col]",
                     incorrect_msg = "Simply print out the result of the slicing, no need to assign it to a new variable.")

success_msg("Great job! Continue to Chapter 2 to learn about tidying data.");
```
