--- 
title_meta  : Chapter 2
title       : Data the Tidy Way
description : "In this chapter, you will learn some of the fundamentals of taking raw data and cleaning it to form tidy data, the fundamental currency of data science and visualization tools such as ggplot2."

--- type:NormalExercise xp:100 skills:1 key:3db54e3447
## Intro to Data Science and Some (Un)Tidy Data

Now that you've been introduced to the fundamentals of programming in R, the next step is to learning how to data science in R. Here, we'll present some of the concepts we wish we would have learned when we first started, and go through some example data on the way. (1, see hint)

By the end of this chapter, you should be familiar with:

- understanding how and why R is useful for data science
- explain the advantages and concepts of tidy data
- familiarity with modern R conventions that make it a powerful tool for data science

On to our first concept: **tidy data**

Many times when we run into data from the wild, it is often a jumble of messy character strings, confounding variables, missing data, and worse. The process of converting a *raw data* to *tidy data* goes by many names - here we will refer to it as *data munging* or *cleaning*. The goal of this process is singular:

> **"A dataset is a collection of *values*, organized in two ways: every value belongs to a *variable* and *observation*.**
> **A *variable* contains all *values* that measure the same underlying attribute across units.**
> **An *observation* contains all *values* measured on the same unit across attributes.**

To emphasize this concept, take a look at the sample code on the right showing untidy data (presented in two different ways), and print the `untidy_df` and `untidy_df_2` object out (hint: Ctrl/Cmd+Enter to run a line!). How does this dataset run counter to the principle stated above? 


*** =instructions
- Take a moment to ponder how you might reorganize this data, then simply click submit to move forward!


*** =hint
(1) These concepts introduced here lean heavily on the formal treatise by Hadley Wickham, the lead developer of many of the packages we will be using throughout this course. See the [tidy data paper](http://vita.had.co.nz/papers/tidy-data.html) and the accompanying [R code heavy version](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html). Definitely take the time to checkout these resources!


*** =pre_exercise_code
```{r}
# run this without showing user
x <- 5
```

*** =sample_code
```{r}
# Untidy data example
untidy_df <- data.frame(name = c("John Smith", "Jane Doe", "Mary Johnson"),
                        treatmenta = c(NA, 4, 6),
                        treatmentb = c(18, 1, 7))
untidy_df

# Untidy data example 2
untidy_df_2 <- data.frame(treatment = c("a", "b"),
                          John.Smith = c(NA, 18),
                          Jane.Doe = c(4, 1),
                          Mary.Johnson = c(6, 7))
untidy_df_2
```

*** =solution
```{r}

```

*** =sct
```{r}
test_error()
test_object("x",
            undefined_msg = "Make sure to define `x`!",
            incorrect_msg = "Have you correctly assigned 5 to `x`!")
success_msg("Good job! Head over to the next exercise!")

```



--- type:NormalExercise lang:r xp: skills: key:489128dafe
## A Tidy Sample

So we just saw what some untidy looks like, with all of its variable and values mixed up. If you took a moment to ponder the format of the previous data (to remind yourself, run the line `untidy_df` and see the output in the console), you took stock of the following.

- There exist three different people with different *names*, who underwent different *treatments*, `treatmenta` and/or `treatmentb`, which resulted in different *values* in each case (with no treatment application (e.g. missing data) represented as `NA`)

Given this, we can come up with three *variables* describing each unique *observation* (application of treatment X to person Y with readout/value Z):

- name
- treatment
- value

Hence, we can rewrite our untidy data into a tidy format! See the sample code for the solution. 

With this in mind, let's move on to a real world dataset 'from the wild'!

*** =instructions
Check out the tidy transformation of our untidy data, and get ready to start coding!

*** =hint
"It is often said that 80% of data analysis is spent on the process of cleaning and preparing
the data (Dasu and Johnson 2003)" - if you haven't already, check out [this paper on tidy data](http://vita.had.co.nz/papers/tidy-data.pdf) for extra credit!

*** =pre_exercise_code
```{r}
# Untidy data example
untidy_df <- data.frame(name = c("John Smith", "Jane Doe", "Mary Johnson"),
                        treatmenta = c(NA, 4, 6),
                        treatmentb = c(18, 1, 7))

```

*** =sample_code
```{r}
untidy_df # run me to see the untidy version

# Note: you cannot run this code line by line
# Instead, select the entire region and then Cmd/Ctrl+Enter and run
tidy_df <- data.frame(name = rep(c("Jane Doe", "John Smith", "Mary Johnson"), each = 2),
                      treatment = rep(c("a", "b"), 3),
                      value = c(4, 1, NA, 18, 6, 7))
tidy_df

```

*** =solution
```{r}

```

*** =sct
```{r}
test_error()
test_object("untidy_df",
            undefined_msg = "Make sure to define `x`!",
            incorrect_msg = "Have you correctly assigned 5 to `x`!")
success_msg("Good job! Head over to the next exercise!")

```




--- type:NormalExercise lang:r xp: skills: key:75bda68f14
## Real (Messy) Data: Income Distribution by Religion

Time for you to take a look at some actual, raw data. Here we have a survey done by Pew Research that examines the relationship between income and religious affiliation. We've already loaded the dataset up for you in the background under `pew`, but if you want to explore it on your own computer, [here is the data](http://stat405.had.co.nz/data/pew.txt).

Recalling chapter 1, check out the dataset on your own, and let's apply our tidy thinking to figure out what we need to change.



*** =instructions
- What do the first and last rows of the data look like?
- What are the dimensions? 
- What does a simple (naive) summary of the data give us? 
- What is the structure of the data?

*** =hint
Use the functions you learned from chapter 1 - namely, `head()`, `tail()`, `dim()`, `summary()`, and `str()` to preview the dataset on hand.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1886/datasets/pew.RData"))

```

*** =sample_code
```{r}
## Your code here
pew

```

*** =solution
```{r}

```

*** =sct
```{r}
test_error()
test_object("pew",
            undefined_msg = "Make sure to define `x`!",
            incorrect_msg = "Have you correctly assigned 5 to `x`!")
success_msg("Good job! Head over to the next exercise!")

```

--- type:NormalExercise lang:r xp: skills: key:426a68171c
## Refactoring Columns Into a Variable

You might have noticed that the columns contain a variety of income *levels* (e.g. *values*)..so you might consider *income* to be a *variable* in its own right. Furthermore, an individual observation will be a combination of a religion crossed with a specific income level.

Before we go on, first, I highly recommend checking out the [Data Wrangling with dplyr and tidyr Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) by RStudio, as this will be an invaluable resource throughout this course (and in your work).

Back to our data..what we want to do is *gather* our columns into a single variable - income. To accomplish this, let's use the `tidyr` package. A quick example with the built-in `iris` dataset is shown in the Sample Code.


*** =instructions
Similar to the `iris` dataset in the sample code, refactor the `pew` dataset such that the data is tidy, with variables `religion`, `income`, and `value` into an object called `pew_df`.

*** =hint
See the sample code and your data wrangling cheat sheet for an example of how to use `tidyr::gather()`. Note that the opposite of this is `tidyr::spread()`.

*** =pre_exercise_code
```{r}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_1886/datasets/pew.RData"))
library(tidyr)

```

*** =sample_code
```{r}
# Example of gathering variables with the iris dataset
# Here, we specify new variables: 
# - a *key* column, "measurement" (Sepal.Length, Petal.Width, etc.)
# - a *value* column, which are the values we measured
# - the *columns* (1:4) where values are in the original data.frame
iris_df <- tidyr::gather(iris, measurement, value, 1:4)
iris_df[1:3, ]


```

*** =solution
```{r}
dat <- tidyr::gather(pew, religion, income)
colnames(dat) <- c("religion", "income", "value")

```

*** =sct
```{r}
test_error()
test_object("pew_df",
            undefined_msg = "Make sure to define `pew_df`!",
            incorrect_msg = "Have you munged `pew`?")
success_msg("Good job! Head over to the next exercise!")

```

--- type:NormalExercise lang:r xp: skills: key:31d35975ad
## <<<TITLE>>> 


*** =instructions

*** =hint

*** =pre_exercise_code
```{r}

```

*** =sample_code
```{r}

```

*** =solution
```{r}

```

*** =sct
```{r}

```
