--- 
title_meta  : Chapter 2
title       : tidyr/dplyr
description : "tidyr/dplyr"

--- type:NormalExercise xp:100 skills:1
## Intro to Data Science and Tidy Data

Now that you've been introduced to the fundamentals of programming in R, the next step is to learning how to data science in R. Here, we'll present some of the concepts we wish we would have learned when we first started, and go through some example data on the way. (1)

By the end of this chapter, you should be familiar with:

- understanding how and why R is useful for data science
- explain the advantages and concepts of tidy data
- familiarity with modern R conventions that make it a powerful tool for data science

On to our first concept: **tidy data**

Many times when we run into data from the wild, it is often a jumble of messy character strings, confounding variables, missing data, and worse. The process of converting a *raw data* to *tidy data* goes by many names - here we will refer to it as *data munging* or *cleaning*. The goal of this process is singular:

> **"A dataset is a collection of *values*, organized in two ways: every value belongs to a *variable* and *observation*.**
> **A *variable* contains all *values* that measure the same underlying attribute across units.**
> **An *observation* contains all *values* measured on the same unit across attributes.**

To emphasize this concept, take a look at the sample code on the right showing untidy data (presented in two different ways), and print the `untidy_df` and `untidy_df_2` object out (hint: Ctrl/Cmd+Enter to run a line!). How does this dataset run counter to the principle stated above? 


*** =instructions
- Take a moment to ponder how you might reorganize this data, then simply click submit to move forward!
 
(1) These concepts introduced here lean heavily on the formal treatise by Hadley Wickham, the lead developer of many of the packages we will be using throughout this course. See the [tidy data paper](http://vita.had.co.nz/papers/tidy-data.html) and the accompanying [R code heavy version](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html). Definitely take the time to checkout these resources!


*** =hint
Why did you click this? You don't need a hint to click submit!

*** =pre_exercise_code
```{r}
# run this without showing user
x <- 5
```

*** =sample_code
```{r}
# Untidy data example
untidy_df <- data.frame(name = c("John Smith", "Jane Doe", "Mary Johnson"),
                        treatmenta = c(NA, 4, 6),
                        treatmentb = c(18, 1, 7))
untidy_df

# Untidy data example 2
untidy_df_2 <- data.frame(treatment = c("a", "b"),
                          John.Smith = c(NA, 18),
                          Jane.Doe = c(4, 1),
                          Mary.Johnson = c(6, 7))
untidy_df_2
```

*** =solution
```{r}

```

*** =sct
```{r}
test_error()
test_object("x",
            undefined_msg = "Make sure to define `x`!",
            incorrect_msg = "Have you correctly assigned 5 to `x`!")
success_msg("Good job! Head over to the next exercise!")

```


